Model Checking Differentially Private Properties

### 2018-01-28 applications.tex
model model checking story

Consider monitoring events of the types $a$ and $b$ continuously.
A sequence of event types is said to be an observation. Two
obsevations are {neighbors} if one can be obtained from the other 
by replacing event types any number of times, that is, if they are of
the same length. %\bw{Later, we will consider a different definition of
%neighborhood by restricting the number of replacements.} 
We would like to inquire the number of event types in
observations. For instance, the query result result is $1$ 
for observations $a \cdots a$ and $b \cdots b$ of arbitrary positive
lengths; and it is $2$ for other non-empty observations.

>>>>

Consider monitoring events of the types $a$ and $b$ continuously.
A sequence of event types is said to be an observation. Two
obsevations are {neighbors} if one can be obtained from the other
by replacing event types any number of times, that is, if they are of
the same length. %\bw{Later, we will consider a different definition of
%neighborhood by restricting the number of replacements.}
We would like to inquire the number of event types in
observations. For instance, the query result is $1$
for observations $a \cdots a$ and $b \cdots b$ of arbitrary positive
lengths; and it is $2$ for other non-empty observations.

### 2018-01-28 mdp.tex
model model checking story

As we describe earlier, the user has the possibility of asking different mechanisms.
We use Markov decision processes to model such
interactive mechanisms. Specifically, external inputs are modeled by
actions. Behaviors associated with different inputs are modeled by
distributions associated with actions.

@@

\subsection{Semantics}
The semantics over Markov chains generalizes to Markov decision
processes as well. Note that two states with different enabled actions
are trivially distinguishable; no privacy can be preserved. Without
loss of generality, we consider reactive Markov decision processes.
Let $M = (S, \Act, \wp, L)$ be a reactive Markov
decision process. Define the satisfaction
relation $M, \neighbor{S}, s \models \Phi$ as follows.
\begin{eqnarray*}
  M, \neighbor{S}, s \models \top\\
  M, \neighbor{S}, s \models p
  & \textmd{ if } &
  p \in L(s)\\
  M, \neighbor{S}, s \models \neg \Phi
  & \textmd{ if } &
  M, \neighbor{S}, s \not\models \Phi\\
  M, \neighbor{S}, s \models \Phi_0 \wedge \Phi_1
  & \textmd{ if } &
  M, \neighbor{S}, s \models \Phi_0 \textmd{ and }
  M, \neighbor{S}, s \models \Phi_1\\
  M, \neighbor{S}, s \models \PJ{J} \phi
  & \textmd{ if } &
  \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \in J
  \textmd{ for every scheduler } \scheduler{S}\\
  M, \neighbor{S}, s \models \dpriv{\epsilon}{\delta} \phi
  & \textmd{ if } &
  \textmd{for all } t \textmd{ with } s \neighbor{S} t \textmd{ and
   scheduler } \scheduler{S}, 
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq 
  e^{\epsilon} \cdot
   \\
  & & 
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
   \textmd{ and }
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq e^{\epsilon} \cdot
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
\end{eqnarray*}

@@

\noindent
\emph{Justification of the scheduler class.}
\lz{related to language equivalence...}

\noindent
\emph{Discussion.}
\lz{this part not needed?}
Another definition of the differentially private operator
might be:
\begin{eqnarray*}
  M, \neighbor{S}, s \models \dpriv{\epsilon}{\delta}^{\mathit{bad}} \phi
  & \textmd{ if } &
  \textmd{for all } t \textmd{ with } s \neighbor{S} t \textmd{ and
  scheduler } \scheduler{S},
  \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
  e^{\epsilon} \cdot\\
  && \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta \textmd{ and }
  \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
  e^{\epsilon} \cdot \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi}
  + \delta
\end{eqnarray*}
A state satisfies $\dpriv{\epsilon}{\delta}^{\mathit{bad}} \phi$ if
no scheduler could differentiate the probabilities of paths satisfying
$\phi$ from neighbors. Such a definition would allow strategies to
take different actions from different states. It might need different
action sequences to distinguish neighbors. Two neighbors could still be
differentiated by a action sequence and lose their privacy. We hence
prefer the original definition.

@@

In general, the computation of the maximal probability for MDPs is non-trivial. \lz{we argue that it is difficult and leave as future work}

Finally, we the property
$\wedge_{i\in N} \dpriv{\epsilon}{0}(\X^k \top) \wedge \dpriv{\epsilon}{0}(F \top)$ that is used to express the privacy for our threshold Markov chains. It is only a valid formula if we would allow infinite conjunctions. However, it is interesting to note the relation to language equivalence for probabilistic automata\lz{to be discussed}

>>>>

As we describe earlier, the user has the possibility of querying multiple mechanisms.
We use Markov decision processes to model such
interactive mechanisms. Specifically, external inputs are modeled by
actions. Behaviors associated with different inputs are modeled by
distributions associated with actions.

@@

\subsection{Semantics}
The semantics over Markov chains generalizes to Markov decision
processes as well. Note that two states with different enabled actions
are trivially distinguishable; no privacy can be preserved. Without
loss of generality, we consider reactive Markov decision processes.
Let $M = (S, \Act, \wp, L)$ be a reactive Markov
decision process. Define the satisfaction
relation $M, \neighbor{S}, s \models \Phi$ for $\PJ{J}$ and $\dpriv{\epsilon}{\delta} \phi$ (others are standard) as follows.
\begin{eqnarray*}
%  M, \neighbor{S}, s \models \top\\
%  M, \neighbor{S}, s \models p  & \textmd{ if } &  p \in L(s)\\
%  M, \neighbor{S}, s \models \neg \Phi  & \textmd{ if } &  M, \neighbor{S}, s \not\models \Phi\\
%  M, \neighbor{S}, s \models \Phi_0 \wedge \Phi_1  & \textmd{ if } &  M, \neighbor{S}, s \models \Phi_0 \textmd{ and }
%  M, \neighbor{S}, s \models \Phi_1\\
  M, \neighbor{S}, s \models \PJ{J} \phi
  & \textmd{ if } &
  \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \in J
  \textmd{ for every scheduler } \scheduler{S}\\
  M, \neighbor{S}, s \models \dpriv{\epsilon}{\delta} \phi
  & \textmd{ if } &
  \textmd{for all } t \textmd{ with } s \neighbor{S} t \textmd{ and
   scheduler } \scheduler{S},
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
  e^{\epsilon} \cdot
   \\
  & &
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
   \textmd{ and }
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq e^{\epsilon} \cdot
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
\end{eqnarray*}

@@

\noindent
\emph{Justification of the scheduler class.}
For MDPs a richer class of scheduler classes have been considered in the literature, see \cite{BK:08:PMC}. It may depend on the full history of state and action sequences, and the decision can be randomised. If we use such scheduler, another definition of the differentially private operator
might be:\lz{put $bad$ on $\models$?}
\begin{eqnarray*}
  M, \neighbor{S}, s \models \dpriv{\epsilon}{\delta}^{\mathit{bad}} \phi
  & \textmd{ if } &
  \textmd{for all } t \textmd{ with } s \neighbor{S} t \textmd{ and
  scheduler } \scheduler{S},
  \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
  e^{\epsilon} \cdot\\
  && \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta \textmd{ and }
  \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
  e^{\epsilon} \cdot \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi}
  + \delta
\end{eqnarray*}
A state satisfies $\dpriv{\epsilon}{\delta}^{\mathit{bad}} \phi$ if
no scheduler could differentiate the probabilities of paths satisfying
$\phi$ from neighbors. Such a definition would allow strategies to
take different actions from different states. It might need different
action sequences to distinguish \lz{or mimic?} neighbors. \lz{add However,?}Two neighbors could still be
differentiated by an action sequence and lose their privacy.

We consider only sequences $\scheduler{S}\in Act^\omega$, which correspond to the query sequences. In the differential privacy context, neighbor states exhibit same action behaviours. And we will see later that 
our semantics agrees with the notion of differential privacy in the literature for such systems.
Wpe hence
prefer the original definition.

@@

In general, the problem is non-trivial. We consider the formula $\phi:=F B$ with $B\subseteq S$ and assume that states in $B$ are absorbing (with only self-loops). For the case $\epsilon=0$, the condition reduces to
$\myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{F B} -
\myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{F B}\le \delta$. If $\delta=0$ it is the classical language equivalence problem for probabilistic automata~\cite{Rabin63}, which can be solved in polynomial time \cite{Tzeng92}. However, if $\delta>0$, the problem
becomes a metric version of the language equivalence problem, see \cite{FengZ13} for the definition. To the best of our knowledge, it has not be investigated in the literature. Thus we leave the general model checking algorithm as future works. \lz{see this}

### 2018-01-28 preliminaries.tex
use standard differential privacy definition e^{\epsilon}
model model checking story

We denote the data universe by $\mathcal{X}$; $x \in \mathcal{X}^n$ is
a database with $n$ rows from the data universe. Two databases $x$ and
$x'$ are \emph{neighbors} (denoted by $d(x, x') \leq 1$) if they are
identical except at most one row. A \emph{query} $f$ is a function
from $\mathcal{X}^n$ to its range $R$. The \emph{sensitivity} of the
query $f$ (written $\Delta (f)$) is $\max_{d(x, x') \leq 1} | f (x) -
f (x') |$. For instance, a \emph{counting} query counts the number
of rows with certain attributes (say, female). The sensitivity of a
counting query is $1$ since any neighbor can change the count by at
most one. We only consider queries with finite  ranges for simplicity.
A \emph{data analysis mechanism} (or
\emph{mechanism} for brevity) $M_f$ for a query $f$
is a randomized algorithm with inputs in $\mathcal{X}^n$ and outputs
in $\tilde{R}$.
A mechanism may not have the same output range as its query, that is,
$\tilde{R} \neq R$ in general.
A mechanism $M_f$ for $f$ is \emph{oblivious} if
$\Pr[M_f(x) = \tilde{r}] = \Pr[M_f(x') = \tilde{r}]$ for every
$r \in \tilde{R}$ when $f (x) = f (x')$. In words, outputs of an
oblivious mechanism depend on the query result $f (x)$ but not on the
input database $x$. The order of rows in a database, for instance, is
irrelevant to oblivious mechanisms. A database $x$ is
\emph{$(\epsilon, \delta)$-close} to $x'$ in a mechanism
$M_f$ if for every $\tilde{r} \in \tilde{R}$,
\[
\Pr[M_f (x) = \tilde{r}] \leq \epsilon \Pr[M_f (x') =
\tilde{r}] + \delta.\footnote{The standard definition requires
$\Pr[M_f (x) = \tilde{r}] \leq e^\epsilon \Pr[M_f (x') =
\tilde{r}] + \delta$~\cite{DMNS:06:CNSPD,D:06:DP}.
Since $\epsilon = e^{\ln \epsilon}$, our $(\epsilon, \delta)$-closedness is
equivalent to the standard $(\ln \epsilon,
\delta)$-closedness~\cite{GRS:09:UUPM,GRS:12:UUPM}.}
\]
A mechanism $M_f$ is \emph{$(\epsilon, \delta)$-differentially
  private} % $(\epsilon, \delta \approx 0)$
if for every $x, x' \in \mathcal{X}^n$ with $d(x, x') \leq 1$,
$x$ is $(\epsilon, \delta)$-close to $x'$ in $M_f$.

@@

Let $\Act$ denote a finite set of \emph{actions}.
A \emph{(finite) Markov decision process} $M = (S, \Act, \wp, L)$ consists
of a \emph{transition probability function} $\wp : S \times \Act
\times S \rightarrow [0, 1]$ with $\sum_{t \in S} \wp(s, \alpha, t)
= 1$ for every $s \in S$ and $\alpha \in \Act$ and $S$,
$L$ as for Markov chains. Note the model we consider is \emph{reactive}\footnote{In the literature, such model is referred to as \emph{probabilistic automata}\cite{}. In the model checking context, the notion MDPs is more popular.} as $\Act(s) = \Act$
for every $s \in S$.
A \emph{path} $\pi$ in $M$ is an infinite sequence $\pi_0 \alpha_1
\pi_1 \cdots \pi_n \alpha_{n+1} \cdots$ with
$\wp(\pi_i, \alpha_{i+1}, \pi_{i+1}) > 0$ for all $i \geq 0$.
Similarly, we write $\pi[j]$ for the suffix $\pi_j \alpha_{j+1}
\pi_{j+1} \cdots$ of $\pi$.
\hide{
For any measurable set $\Pi$ of paths, we
write $\Pr[\Pi]$ for the \emph{probability measure} of $\Pi$.
}

Let $M = (S, \Act, \wp, L)$ be a Markov decision process. A
\emph{scheduler} for $M$ is simply an infinite sequence  $\scheduler{S}\in \Act^\omega$.
A path $\pi =
\pi_0 \alpha_1 \pi_1 \cdots \pi_n \alpha_{n+1} \cdots$ is an
\emph{$\scheduler{S}$-path} if $\alpha_{i+1} = \scheduler{S}(\pi_0
\pi_1 \cdots \pi_i)$ for all $i \geq 0$.
Note that a Markov decision process with a
scheduler $\scheduler{S}$ induces a Markov chain $M_{\scheduler{S}} =
(S^+, \wp_{\scheduler{S}}, L')$ where $L' (\sigma s) = L (s)$,
$\wp_{\scheduler{S}} (\sigma s, \sigma s t) = \wp (s,
\scheduler{S}(\sigma s), t)$ for $\sigma \in S^*$ and $s, t \in S$.
\hide{
Subsequently, we write
$\Pr_{\scheduler{S}} (\Pi)$ for the \emph{probability measure} of the
measurable path set $\Pi$ on $M$ given the scheduler $\scheduler{S}$.
}

>>>>

We denote the data universe by $\mathcal{X}$; $x \in \mathcal{X}^n$ is
a database with $n$ rows from the data universe. Two databases $x$ and
$x'$ are \emph{neighbors} (denoted by $d(x, x') \leq 1$) if they are
identical except at most one row. A \emph{query} $f$ is a function
from $\mathcal{X}^n$ to its range $R$. The \emph{sensitivity} of the
query $f$ (written $\Delta (f)$) is $\max_{d(x, x') \leq 1} | f (x) -
f (x') |$. For instance, a \emph{counting} query counts the number
of rows with certain attributes (say, female). The sensitivity of a
counting query is $1$ since any neighbor can change the count by at
most one. We only consider queries with finite  ranges for simplicity.
A \emph{data analysis mechanism} (or
\emph{mechanism} for brevity) $M_f$ for a query $f$
is a randomized algorithm with inputs in $\mathcal{X}^n$ and outputs
in $\tilde{R}$.
A mechanism may not have the same output range as its query, that is,
$\tilde{R} \neq R$ in general.
A mechanism $M_f$ for $f$ is \emph{oblivious} if
$\Pr[M_f(x) = \tilde{r}] = \Pr[M_f(x') = \tilde{r}]$ for every
$r \in \tilde{R}$ when $f (x) = f (x')$. In words, outputs of an
oblivious mechanism depend on the query result $f (x)$ but not on the
input database $x$. The order of rows in a database, for instance, is
irrelevant to oblivious mechanisms. A database $x$ is
\emph{$(\epsilon, \delta)$-close} to $x'$ in a mechanism
$M_f$ if for every $\tilde{r} \in \tilde{R}$,
\[
\Pr[M_f (x) = \tilde{r}] \leq e^{\epsilon} \Pr[M_f (x') =
\tilde{r}] + \delta.
%\footnote{The standard definition requires
%$\Pr[M_f (x) = \tilde{r}] \leq e^\epsilon \Pr[M_f (x') =
%\tilde{r}] + \delta$~\cite{DMNS:06:CNSPD,D:06:DP}.
%Since $\epsilon = e^{\ln \epsilon}$, our $(\epsilon, \delta)$-closedness is
%equivalent to the standard $(\ln \epsilon,
%\delta)$-closedness~\cite{GRS:09:UUPM,GRS:12:UUPM}.}
\]
A mechanism $M_f$ is \emph{$(\epsilon, \delta)$-differentially
  private} % $(\epsilon, \delta \approx 0)$
if for every $x, x' \in \mathcal{X}^n$ with $d(x, x') \leq 1$,
$x$ is $(\epsilon, \delta)$-close to $x'$ in $M_f$.

@@

A \emph{(finite) Markov decision process} 
(MDP)~\footnote{The MDP we consider is \emph{reactive} in the sense that 
all actions are enabled in every state. If equipped with a set of final states $F\subseteq S$, such model is
referred to as \emph{probabilistic automaton} in the literature~\cite{Rabin63}, and language equivalence/inclusion
problems have been investigated. In the model checking context, the notion MDPs is mostly used.}
 $M = (S, \Act, \wp, L)$ consists
of 
a finite set of \emph{actions} $\Act$,
a \emph{transition probability function} $\wp : S \times \Act
\times S \rightarrow [0, 1]$ with $\sum_{t \in S} \wp(s, \alpha, t)
= 1$ for every $s \in S$ and $\alpha \in \Act$ and $S$,
$L$ as for Markov chains.
A \emph{path} $\pi$ in $M$ is an infinite sequence $\pi_0 \alpha_1
\pi_1 \cdots \pi_n \alpha_{n+1} \cdots$ with
$\wp(\pi_i, \alpha_{i+1}, \pi_{i+1}) > 0$ for all $i \geq 0$.
Similarly, we write $\pi[j]$ for the suffix $\pi_j \alpha_{j+1}
\pi_{j+1} \cdots$ of $\pi$.
\hide{
For any measurable set $\Pi$ of paths, we
write $\Pr[\Pi]$ for the \emph{probability measure} of $\Pi$.
}

Let $M = (S, \Act, \wp, L)$ be an MDP. A
\emph{scheduler} for $M$ is simply an infinite sequence  $\scheduler{S}\in \Act^\omega$.
A path $\pi =
\pi_0 \alpha_1 \pi_1 \cdots \pi_n \alpha_{n+1} \cdots$ is an
\emph{$\scheduler{S}$-path} if $\alpha_{i+1} = \scheduler{S}(\pi_0
\pi_1 \cdots \pi_i)$ for all $i \geq 0$.
Note that an MDP with a
scheduler $\scheduler{S}$ induces a Markov chain $M_{\scheduler{S}} =
(S^+, \wp_{\scheduler{S}}, L')$ where $L' (\sigma s) = L (s)$,
$\wp_{\scheduler{S}} (\sigma s, \sigma s t) = \wp (s,
\scheduler{S}(\sigma s), t)$ for $\sigma \in S^*$ and $s, t \in S$.
\hide{
Subsequently, we write
$\Pr_{\scheduler{S}} (\Pi)$ for the \emph{probability measure} of the
measurable path set $\Pi$ on $M$ given the scheduler $\scheduler{S}$.
}

### 2018-01-28 specifications.tex
model model checking story

 Define the $\dpCTL$ formula
$\psi = \dpriv{\ln 2}{0} (\X out_0) \wedge \dpriv{\ln 2}{0} (\X out_1) \wedge
\cdots \wedge \dpriv{\ln 2}{0} (\X out_5)$. If the state $in_k$ satisfies
$\psi$ for $k = 0, \ldots, 5$, then the $.5$-geometric mechanism is
$(\ln 2, 0)$-differentially private.

@@

Now we consider the formula $\dpriv{\epsilon_2}{\delta_2} (\X (Y\wedge \X Y))$ under the model in Figure~\ref{figure:2-dp-mc}. Applying the compositional theorem, one can choose $\epsilon_2=2\epsilon_1=0$ and $\delta_2=2\delta_1=1$. However, we can check easily that one gets better privacy parameter $(0,0.5)$ or $(-\infty,\frac{9}{16})$. \bw{use $-\infty$ instead of $\ln 0$. is it okay?}

@@

>>>>

 Define the formula
$\psi = \dpriv{\ln 2}{0} (\X out_0) \wedge \dpriv{\ln 2}{0} (\X out_1) \wedge
\cdots \wedge \dpriv{\ln 2}{0} (\X out_5)$. If the state $in_k$ satisfies
$\psi$ for $k = 0, \ldots, 5$, then the $.5$-geometric mechanism is
$(\ln 2, 0)$-differentially private.

@@

Now we consider the formula $\dpriv{\epsilon_2}{\delta_2} (\X (Y\wedge \X Y))$ under the model in Figure~\ref{figure:2-dp-mc}. Applying the compositional theorem, one can choose $\epsilon_2=2\epsilon_1=0$ and $\delta_2=2\delta_1=1$. However, we can check easily that one gets better privacy parameter $(0,0.5)$ or $(-\infty,\frac{9}{16})$. \bw{use $-\infty$ instead of $\ln 0$. is it okay?}\lz{lets just remove it ;)}

@@

Finally, we the property
$\wedge_{i\in N} \dpriv{\epsilon}{0}(\X^k \top) \wedge \dpriv{\epsilon}{0}(F \top)$ that is used to express the privacy for our threshold Markov chains. It is only a valid formula if we would allow infinite conjunctions. However, it is interesting to note the relation to language equivalence for probabilistic automata\lz{to be discussed}
